---
title: "Anaylsis of Impact of Major Storms and Weather Events"
author: "Zach Valadez"
date: "July 1, 2018"
output: html_document
---

## Sympnosis
The goal of this anaylsis is to exlore the effect of severe storms and weather events in an area by using
NOAA storm database, which covers events from 1950 and end in November 2011.  
We will look at how these storms and weather events have effected  
1. Fatalities and injuries  
2. Property and crops damage
[NOAA storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

## Data Processing
### Loading Data
We will first download the data we want to observe then make it into a data table to make it easier
to work with.
```{r}
library(data.table)
library(ggplot2)

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(url, paste0(getwd(),"/repdata%2Fdata%2FStormData.csv.bz2"))
stormDF <- read.csv(paste0(getwd(),"/repdata%2Fdata%2FStormData.csv.bz2"))

stormDT <- as.data.table(stormDF)
```
### Looking at Column Names
```{r}
colnames(stormDT)
```
### Subsetting Data
We now are going to subset the data and take out only what we are interested in.
```{r}
colsRemoved <- colnames(stormDT[, !c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP", 
                                     "CROPDMG", "CROPDMGEXP")])

stormDT[, c(colsRemoved) := NULL]

stormDT <- stormDT[(EVTYPE != "?" & (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), 
                   c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP")]

```

### Converting Exponents Columns into Actual Exponents
Right now in our data table our exponent column is made up on different characters and not real expontents.
This is true for exponent columns for PROPDMGEXP and CROPDMGEXP. This will help these columns look cleaner
as well as to be used to calculate property crop damage cost.
```{r}
propDmgKey <- c("\"\"" = 10^0, "-" = 10^0, "+" = 10^0,"0" = 10^0,"1" = 10^1,"2" = 10^2,"3" = 10^3,
                "4" = 10^4,"5" = 10^5,"6" = 10^6,"7" = 10^7,"8" = 10^8,"9" = 10^9,"H" = 10^2,"K" = 10^3,
                 "M" = 10^6,"B" = 10^9)

cropDmgKey <- c("\"\"" = 10^0,"?" = 10^0, "0" = 10^0,"K" = 10^3,"M" = 10^6,"B" = 10^9)

stormDT[, PROPDMGEXP := propDmgKey[as.character(stormDT[,PROPDMGEXP])]]
stormDT[is.na(PROPDMGEXP), PROPDMGEXP := 10^0]

stormDT[, CROPDMGEXP := cropDmgKey[as.character(stormDT[,CROPDMGEXP])]]
stormDT[is.na(CROPDMGEXP), CROPDMGEXP := 10^0]
```
### Creating Cost Columns for Property and Crops
```{r}
stormDT <- stormDT[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP,
                       CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```

### Total Property and Crop Damage Cost
```{r}
totalCostDT <- stormDT[, .(propCost = sum(propCost), cropCost = sum(cropCost), 
                         total_cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]
totalCostDT <- totalCostDT[order(-total_cost),]
totalCostDT <- totalCostDT[1:10, ]
head(totalCostDT, 5)
```

### Total Number of FATALITIES and Injuries
```{r}
totalInjuriesDT <- stormDT[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), 
                         total_injuries = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]
totalInjuriesDT <- totalInjuriesDT[order(-total_injuries),]
totalInjuriesDT <- totalInjuriesDT[1:10, ]
head(totalInjuriesDT, 5)
```

## Results
### Events that caused the most injuries
First we will create a data table to help create a plot
```{r}
harmfulEvent <- melt(totalInjuriesDT, id.vars = "EVTYPE", variable.name = "harmfulThing")
head(harmfulEvent, 5)
```

Now we are going to make our plot.
```{r}
healthPlot <- ggplot(harmfulEvent, aes(x = reorder(EVTYPE, -value), y = value))
healthPlot = healthPlot + geom_bar(stat = "identity", aes(fill = harmfulThing), position = "dodge") +
    labs(title = "Most Harmful Storms in the US", xlabs = "Event Type", ylabs = "Frequency Count") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), plot.title = element_text(hjust = 0.5))
healthPlot
```

### Events that caused the most damage
First we will create a data table to help create a plot once again
```{r}
destructiveEvent <- melt(totalCostDT, id.vars = "EVTYPE", variable.name = "destructiveThing")
head(destructiveEvent, 5)
```

Now we are going to make our plot.
```{r}
costPlot <- ggplot(destructiveEvent, aes(x = reorder(EVTYPE, -value), y = value))
costPlot = costPlot + geom_bar(stat = "identity", aes(fill = destructiveThing), position = "dodge") +
    labs(title = "Most Destructive Storms in the US", xlabs = "Event Type", ylabs = "Cost (dollars)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), plot.title = element_text(hjust = 0.5))
costPlot
```


